The Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-09-28 paper.tex
Revisao Secao 2.3

The introductory example of Figure~\ref{lst.abro} illustrates how synchronous 
languages can abort awaiting lines of execution without tweaking them with 
synchronization primitives.
In contrast, traditional (asynchronous) multi-threaded languages cannot express 
thread termination safely~\cite{esterel.preemption,sync_async.threadsstop}.
However, handling abortion when dealing with external resources can be 
challenging, given that external entities are not subject to the same 
synchronous execution discipline.

\begin{figure}
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void STOP, RETRANSMIT, SENDACK;
par/or do
    await STOP;
with
    loop do
        par/or do
            await RETRANSMIT;
        with
            par/and do
                await 1min;
            with
                var _pkt_t buffer;
                <fill-buffer-info>
                _send_enqueue(&buffer);
                await SENDACK;
            end
        end
    end
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.41\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em,firstnumber=11]
<...>
    var _pkt_t buffer;
    <fill-buffer-info>
    finalize
        _send_enqueue(&buffer);
    with
        _send_cancel(&buffer);
    end
    await SENDACK;
<...>
\end{lstlisting}
\end{minipage}
%\rule{8.5cm}{0.37pt}
\caption{The unsafe network protocol in the left is extended with a 
finalization clause in the right to handle abortion properly.
%\caption{ Parallel compositions can describe complex state machines.
\label{lst.fin}
}
\end{figure}

To illustrate threats related to abortion, consider the unsafe example in the 
left of Figure~\ref{lst.fin}, describing the state machine of a data collection 
protocol for sensor networks~\cite{wsn.ctp,ceu.sensys13}.
%
The input events \code{STOP}, \code{RETRANSMIT}, and \code{SENDACK} (line 1) 
represent the external interface of the protocol with a client application.
%
The protocol has to transmit a packet every minute, unless it receives a 
\code{RETRANSMIT} request to immediately re-transmit, or a \code{STOP} request 
to terminate.
%
The protocol is composed of two main trails:
one simply monitors the stopping event (line 3);
the other periodically transmits a packet (lines 5--18).
%
The periodic transmission is a loop that starts two other trails (lines 6--17):
one handles the immediate retransmission request (line 7);
the other transmits the packet%
\footnote{
The underline prefix marks (e.g., \code{\_send\_enqueue}) make interactions 
with external \emph{C} libraries explicit and are required in \CEU.
}
and waits for a confirmation (lines 9--16).
%
%
The actual transmission (lines 12--15) is enclosed with a \code{par/and} that 
takes at least one minute before looping, to avoid flooding the network with 
packets.
Note that the call to \code{\_send\_enqueue} is \emph{asynchronous}, handing to 
the radio driver a pointer to the lexically-scoped packet.
The driver will make the transmission in the background, holding the packet 
until it signals the application with the \code{SENDACK} to acknowledge 
completion.
%
At any time, the client may request a retransmission (line 7), which terminates 
the \code{par/or} (line 6), aborts the ongoing transmission (line 14, if not 
idle), and restarts the loop (line 5).
%
The client may also request to stop the whole protocol at any time (line 3).
%
In the meantime, if the sending trail is be aborted by the \code{STOP} or 
\code{RETRANSMIT} requests, the packet buffer goes out of scope, leaving behind 
a \emph{dangling pointer} in the radio driver, which will possibly transmit 
wrong data.
%
%\subsubsection{Finalization}
%\label{sec.ceu.fin}

The \CEU compiler tracks the interaction of \code{par/or} compositions with 
local variables and stateful \emph{C} functions (e.g., device drivers) in order 
to preserve safe abortion of trails~\cite{ceu.sensys13,ceu.mod15}.
%
%Finalization clauses are fundamental to preserve the orthogonality of 
%\code{par/or} compositions in SSRP.
%
For instance, \CEU refuses to compile the program in the left of 
Figure~\ref{lst.fin}, enforcing the programmer to write a \emph{finalization} 
clause to accompany the stateful \emph{C} call.
The code in the right of Figure~\ref{lst.fin} properly cancels the packet 
transmission when the block of \code{buffer} goes out of scope, i.e., the 
finalization clause (after the \code{with}) executes automatically on external 
abortion.%
\footnote{
Note that the compiler only enforces the programmer to write the finalization 
clause, but cannot check if it actually handles the resource properly.
}
%
%Note also that \CEU environments rely on \emph{C} libraries that only provide 
%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.

>>>>

The introductory example of Figure~\ref{lst.abro} illustrates how synchronous 
languages can abort awaiting lines of execution without tweaking them with 
synchronization primitives.
In contrast, traditional (asynchronous) multi-threaded languages cannot express 
thread termination safely~\cite{esterel.preemption,sync_async.threadsstop}.
Still, handling abortion when dealing with external resources is challenging 
because they are not subject to the same synchronous execution discipline.

\begin{figure}
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void STOP, RETRANSMIT, SENDACK;
par/or do
    await STOP;
with
    loop do
        par/or do
            await RETRANSMIT;
        with
            par/and do
                await 1min;
            with
                var _pkt_t buffer;
                <fill-buffer-info>
                _send_enqueue(&buffer);
                await SENDACK;
            end
        end
    end
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.41\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em,firstnumber=11]
<...>
    var _pkt_t buffer;
    <fill-buffer-info>
    finalize
        _send_enqueue(&buffer);
    with
        _send_cancel(&buffer);
    end
    await SENDACK;
<...>
\end{lstlisting}
\end{minipage}
%\rule{8.5cm}{0.37pt}
\caption{The unsafe network protocol in the left, which does not compile, is 
extended with a finalization clause in the right to handle abortion properly.
%\caption{ Parallel compositions can describe complex state machines.
\label{lst.fin}
}
\end{figure}

To illustrate threats related to abortion, consider the unsafe example, which 
does not compile, in the left of Figure~\ref{lst.fin}, describing the state 
machine of a data collection protocol for sensor 
networks~\cite{wsn.ctp,ceu.sensys13}.
%
The input events \code{STOP}, \code{RETRANSMIT}, and \code{SENDACK} (line 1) 
represent the external interface of the protocol.
%
The protocol has to transmit a packet every minute, unless it receives a 
\code{RETRANSMIT} request to immediately re-transmit it, or a \code{STOP} 
request to terminate.
%
The protocol is composed of two main trails:
one simply monitors the stopping event (line 3);
the other periodically transmits the packet (lines 5--18).
%
The periodic transmission is a loop that starts two other trails (lines 6--17):
one handles the immediate retransmission request (line 7);
the other transmits the packet%
\footnote{
The underline prefix marks (e.g., \code{\_send\_enqueue}) make interactions 
with external \emph{C} libraries explicit and are required in \CEU.
}
and waits for a confirmation (lines 9--16).
%
%
The actual transmission (lines 12--15) is enclosed with a \code{par/and} that 
takes at least one minute before looping, in accordance with the specification.
Note that the call to \code{\_send\_enqueue} is \emph{asynchronous}, handing to 
the radio driver a pointer to the lexically-scoped packet.
The driver makes the transmission in the background, holding the packet until 
it signals the application with the \code{SENDACK} to acknowledge completion.
%
At any time, the client may request a retransmission (line 7), which terminates 
the \code{par/or} (line 6), aborts the ongoing transmission (line 14, if not 
idle), and restarts the loop (line 5).
%
The client may also request to stop the whole protocol at any time (line 3).
%
Therefore, if the sending trail is aborted by the \code{STOP} or 
\code{RETRANSMIT} requests, the packet buffer goes out of scope, leaving behind 
a \emph{dangling pointer} in the radio driver, which will possibly transmit 
corrupted data.
%
%\subsubsection{Finalization}
%\label{sec.ceu.fin}

The unsafe example in the left of Figure~\ref{lst.fin} does not compile because 
\CEU tracks the interaction of \code{par/or} compositions with local variables 
and stateful \emph{C} functions (e.g., device drivers) in order to preserve 
safe abortion of trails~\cite{ceu.sensys13,ceu.mod15}.
%
%Finalization clauses are fundamental to preserve the orthogonality of 
%\code{par/or} compositions in SSRP.
%
In fact, \CEU enforces the programmer to write a \emph{finalization} clause to 
accompany the stateful \emph{C} call.
The code in the right of Figure~\ref{lst.fin} properly cancels the packet 
transmission when the block of \code{buffer} goes out of scope, i.e., the 
finalization clause (after the \code{with}) executes automatically on external 
abortion.%
\footnote{
Note that the compiler only enforces the programmer to write the finalization 
clause, but cannot check if it actually handles the resource properly.
}
%
%Note also that \CEU environments rely on \emph{C} libraries that only provide 
%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.
